### 2014-01-22 brains-introduction.tex
Phrasings and obvious mistakes

A third approach, ``cultural epidemiology'' initiated by \citet{sper:expl} recently attracted significant attention. Works such as \citet{Atran03} suggest that this stance could be anthropologically more satisfying than memetics; see \citet{Kuper00}, and \citet{Bloch00}, for further details on the main issues in this debate.
It focuses on the notion of \emph{representation}, linking the cognitive science concept of \emph{mental representation} to the concept of \emph{public representation}, the latter being the equivalent of the former outside of the brain (\hbox{i.e.} in cultural artefacts: texts, utterances, etc.).\footnote{\citet{sper:expl} emphasizes this distinction in his seminal work:
%\begin{quotation}
{%Une représentation peut exister à l'intérieur même de l'utilisateur : il s'agit alors d'une \emph{représentation mentale}. Un souvenir, une hypothèse, une intention sont des exemples de représentations mentales. L'utilisateur et le producteur d'une représentation mentale ne font qu'un. Une représentation peut aussi exister dans l'environnement de l'utilisateur comme par exemple le texte qui est sous vos yeux. Il s'agit alors d'une \emph{représentation publique}. Une représentation publique est généralement un moyen de communication entre un producteur et un utilisateur distincts l'un de l'autre.
``%
A representation may exist inside its user: it is then a \emph{mental representation}, such as a memory, a belief, or an intention. The producer and the user of a mental representation are one and the same person.  A representation may also exist in the environment of its user, as is the case, for instance, of the text you are presently reading: it is then a \emph{public representation}.}%~(\citealp[p.~49]{sper:expl}.)
''
%\end{quotation}
}
The key point here is that representations are not being replicated through a hi-fidelity copying process, but are being reformulated through an interpretation process, and thus evolve. Cultural epidemiology postulates this type of conceptual mutation and suggests that it can be appraised through the notion of ``cultural attractor'', seen as the attraction domains of an underlying semantic dynamical system.
Despite some recent modeling attempts (for instance \citet{Claidiere07}), the development of quantitative measurements focused on the key notion of cultural attractor has remained a relatively hard task and, to our knowledge, this hypothesis has not been empirically validated.

@@

The last decade has however witnessed an avalanche of observable \emph{in vivo} data stemming from online interactions. These large amounts of textual traces of information production by humans are plausibly not records of strictly ``physical'' inter-individual interactions (in the sense of ``real life'' interactions), yet they actually constitute a wealth of observations on the dynamics of public representations. As such, they are prone to radically improve the prospects of empirical validation of the individual-level processes of cultural evolution.

We aim to empirically describe the transformation of a specific type of public representations by focusing on the possible cognitive alterations performed by individuals during reformulation.  % On s'est proposé d'étudier de façon \emph{empirique} la transformation des représentations publiques, en se focalisant sur la mise en évidence d'un biais cognitif dans le traitement  par le cerveau de certaines représentations culturelles. 
To deal with robust and simple cultural representations, we paid attention to the evolution of quotations. Admittedly, these verbatim public representations should not suffer any reformulation alteration (as opposed to more sophisticated expressions and opinions), but still do.  We will in particular exhibit a non-trivial cognitive alteration by which some portions of the quotation, namely individual words, may be modified; and will wonder which are generally the semantic and structural characteristics of these words and their substitutions. 
%On verra que par cette analyse de la reproduction des citations par les auteurs on observe un biais cognitif non trivial ; on observera ainsi un lien entre l'attractivité de certains mots (en termes d'épidémiologie culturelle) et les propriétés sémantiques de ces mêmes mots, calculées sur la base du réseau formé par ces mots.
%On s'est proposé d'étudier de façon \emph{empirique} la transformation des représentations publiques, c'est-à-dire d'explorer à l'aide de données obtenues in vivo l'effet du cerveau sur l'évolution des représentations ; une telle étude permettrait d'observer empiriquement l'effet d'un attracteur culturel tel que défini par Sperber. La question est donc la suivante : peut-on observer un biais cognitif dans le traitement de certaines représentations culturelles par le cerveau, et si oui quel est-il ?On verra que par cette analyse de la reproduction des citations par les auteurs on observe un biais cognitif non trivial ; on observera ainsi un lien entre l'attractivité de certains mots (en termes d'épidémiologie culturelle) et les propriétés sémantiques de ces mêmes mots, calculées sur la base du réseau formé par ces mots.
Broadly, we contend that using this type of data is both equivalent to a large-scale psycholinguistic experiment and at the same time constitutes the first step towards the design of empirically realistic models of cultural evolution.

@@

On the other hand, the study of the \emph{transformation} of public representations has emerged only recently. For one, the notion of ``cultural attractor'' has been the focus of models rendering the evolution of some representations and behaviors not until a few years ago \citep{Claidiere07}. Among the empirical approaches on the mutation of representations, some of the most relevant studies to date consist of a series of papers investigation \emph{quotation} transformations from a large corpus of US blog posts, initially collected and studied by \citet{Leskovec09} and further analyzed by \citet{Simmons11} and \citet{omod-mult}. They generally show several types of regularities and propose models of diffusion-transformation of quotations, which may nonetheless appear to be relatively simplistic from a cognitive viewpoint.  One of the main conclusions of these works lies in the fact that even for the supposedly most stable type of public representation, namely quotations, it is indeed still possible to observe and measure significant transformations.  However, these studies address transformations by focusing on the properties of the author (e.g. news vs. blogs) of the quotation, or the surrounding public space (e.g. quote frequency in the corpus), rather than the very cognitive features which may trigger or, at least, bias these transformations. % Cependant ces auteurs n'étudient les transformations observées qu'en fonction du type d'auteur qui reproduit les citations, et pas en fonction des citations elles-mêmes (en particulier les questions d'ordre cognitif concernant la raison pour laquelle ces transformations ont lieu, ou la raison pour laquelle elles ont lieu dans tel sens et pas dans tel autre, ne sont pas abordées). %Les possibilités offertes par cette étude étant vastes, on a décidé de travailler sur les mêmes données que \citeauthor{Simmons11} pour traiter notre question.

@@

\TB{The study of network-based properties, while heavily discussed around the notion of connectionism and its normative processual models, appeared much more recently as an empirical investigation field \CN} \rk{C'est sûr ? e.g. http://tiny.cc/bl7usw}. Let us mention two interesting and recent studies on that matter, which demonstrate in a strictly \emph{in vitro} framework and at the vocabulary level that word properties computed on a word network are important factors in the cognitive processes and reproduction of those words. %Les deux autres travaux viennent de la psycholinguistique, et posent les fondations pour l'étude des propriétés des représentations (qui seront ici des citations). 
First, \citet{Griffiths07} analyze a task where patients are asked to speak out the first word which comes to their mind when they are presented with a random alphabet letter. The authors show that there exists a link between the ease of recall of words and one of their semantic features, namely their authority position (PageRank) in a language-wide semantic network built from external word association data. % property concernant une tâche où l'on présente aux sujets une lettre de l'alphabet en leur demandant de dire le premier mot commençant par cette lettre qui leur vient à l'esprit. Les auteurs montrent qu'une certaine propriété des mots (le PageRank calculé sur un réseau sémantique construit à partir de données d'associations de mots) est un bon facteur prédictif de quel mot sera rappelé parmi les mots commençant par la lettre demandée ; ils montrent ainsi un lien entre une caractéristique sémantique quantitative des mots et la facilité de rappel de ces mots dans cette tâche. 
A second psycholinguistic study by \citet{Chan10} shows, in a task of picture naming, that words come quicker when they have a higher clustering coefficient in an underlying phonological network (which, again, is  defined from external phonological data). %, étudie une tâche où l'on demande aux sujets de nommer oralement une image. Les auteurs montrent qu'une autre propriété des mots, phonologique cette fois (le coefficient de regroupement des mots calculé sur le réseau phonologique formé par ces mots), a un effet sur la rapidité à nommer l'image présentée : c'est un autre cas de lien entre une caractéristique quantitative (phonologique) d'un mot et la facilité à produire ce mot.

>>>>

A third approach, ``cultural epidemiology'' initiated by \citet{sper:expl} recently attracted significant attention. Works such as \citet{Atran03} suggest that this program is anthropologically more accurate than memetics; see \citet{Kuper00}, and \citet{Bloch00}, for further details on the main issues in this debate.
It focuses on the notion of \emph{representation}, linking the cognitive science concept of \emph{mental representation} to the concept of \emph{public representation}, the latter being the equivalent of the former outside of the brain (\hbox{i.e.} in cultural artefacts: texts, utterances, etc.).\footnote{\citet{sper:expl} emphasizes this distinction in his seminal work:
%\begin{quotation}
{%Une représentation peut exister à l'intérieur même de l'utilisateur : il s'agit alors d'une \emph{représentation mentale}. Un souvenir, une hypothèse, une intention sont des exemples de représentations mentales. L'utilisateur et le producteur d'une représentation mentale ne font qu'un. Une représentation peut aussi exister dans l'environnement de l'utilisateur comme par exemple le texte qui est sous vos yeux. Il s'agit alors d'une \emph{représentation publique}. Une représentation publique est généralement un moyen de communication entre un producteur et un utilisateur distincts l'un de l'autre.
``%
A representation may exist inside its user: it is then a \emph{mental representation}, such as a memory, a belief, or an intention. The producer and the user of a mental representation are one and the same person.  A representation may also exist in the environment of its user, as is the case, for instance, of the text you are presently reading: it is then a \emph{public representation}.}%~(\citealp[p.~49]{sper:expl}.)
''
%\end{quotation}
}
The key point here is that representations are not being replicated through a hi-fidelity copying process, but are being reformulated through an interpretation process, and thus evolve. Cultural epidemiology postulates this type of conceptual mutation and suggests that it can be appraised through the notion of ``cultural attractor'', seen as the attraction domains of an underlying semantic dynamical system.
Despite some recent modeling attempts (for instance \citet{Claidiere07}), the development of quantitative measurements focused on the key notion of cultural attractor has remained a relatively hard task and, to our knowledge, this hypothesis has not been empirically validated.

@@

The last decade has however witnessed an avalanche of observable \emph{in vivo} data stemming from online interactions. While they are not records of strictly ``physical'' inter-individual interactions (in the sense of ``real life'' interactions), these information trails and productions actually constitute a wealth of observations on the dynamics of public representations. As such, they are prone to radically improve the prospects of empirical validation of the individual-level processes of cultural evolution.

We aim to empirically describe the transformation of a specific type of public representations by focusing on the possible cognitive alterations performed by individuals during reformulation.  % On s'est proposé d'étudier de façon \emph{empirique} la transformation des représentations publiques, en se focalisant sur la mise en évidence d'un biais cognitif dans le traitement  par le cerveau de certaines représentations culturelles. 
To deal with robust and simple cultural representations, we paid attention to the evolution of quotations. Although these verbatim public representations should not suffer any alterations as they are repeated (as opposed to more elaborate expressions and opinions), in practice they are oftentimes transformed.  We will in particular exhibit a non-trivial cognitive alteration by which some portions of the quotation, namely individual words, may be modified; and will wonder which are generally the semantic and structural characteristics of these words and their substitutions. 
%On verra que par cette analyse de la reproduction des citations par les auteurs on observe un biais cognitif non trivial ; on observera ainsi un lien entre l'attractivité de certains mots (en termes d'épidémiologie culturelle) et les propriétés sémantiques de ces mêmes mots, calculées sur la base du réseau formé par ces mots.
%On s'est proposé d'étudier de façon \emph{empirique} la transformation des représentations publiques, c'est-à-dire d'explorer à l'aide de données obtenues in vivo l'effet du cerveau sur l'évolution des représentations ; une telle étude permettrait d'observer empiriquement l'effet d'un attracteur culturel tel que défini par Sperber. La question est donc la suivante : peut-on observer un biais cognitif dans le traitement de certaines représentations culturelles par le cerveau, et si oui quel est-il ?On verra que par cette analyse de la reproduction des citations par les auteurs on observe un biais cognitif non trivial ; on observera ainsi un lien entre l'attractivité de certains mots (en termes d'épidémiologie culturelle) et les propriétés sémantiques de ces mêmes mots, calculées sur la base du réseau formé par ces mots.
Broadly, we contend that using this type of data is both equivalent to a large-scale psycholinguistic experiment and at the same time constitutes the first step towards the design of empirically realistic models of cultural evolution.

@@

On the other hand, the study of the \emph{transformation} of public representations has emerged only recently. For one, the notion of ``cultural attractor'' has been the focus of models rendering the evolution of some representations and behaviors not until a few years ago \citep{Claidiere07}. Among the empirical approaches on the mutation of representations, some of the most relevant studies to date consist of a series of papers investigating \emph{quotation} transformations from a large corpus of US blog posts, initially collected and studied by \citet{Leskovec09} and further analyzed by \citet{Simmons11} and \citet{omod-mult}. They generally show several types of regularities and propose models of diffusion-transformation of quotations, which may nonetheless appear to be relatively simplistic from a cognitive viewpoint.  One of the main conclusions of these works lies in the fact that even for the supposedly most stable type of public representation, namely quotations, it is indeed still possible to observe and measure significant transformations.  However, these studies address transformations by focusing on the properties of the author (e.g. news vs. blogs) of the quotation, or the surrounding public space (e.g. quote frequency in the corpus), rather than the very cognitive features which may trigger or, at least, bias these transformations. % Cependant ces auteurs n'étudient les transformations observées qu'en fonction du type d'auteur qui reproduit les citations, et pas en fonction des citations elles-mêmes (en particulier les questions d'ordre cognitif concernant la raison pour laquelle ces transformations ont lieu, ou la raison pour laquelle elles ont lieu dans tel sens et pas dans tel autre, ne sont pas abordées). %Les possibilités offertes par cette étude étant vastes, on a décidé de travailler sur les mêmes données que \citeauthor{Simmons11} pour traiter notre question.

@@

The study of network-based properties, while heavily discussed around the notion of connectionism and its normative processual models, appeared much more recently as an empirical investigation field \CN. Let us mention two interesting and recent studies on that matter, which demonstrate in a strictly \emph{in vitro} framework and at the vocabulary level that word properties computed on a word network are important factors in the cognitive processes and reproduction of those words. %Les deux autres travaux viennent de la psycholinguistique, et posent les fondations pour l'étude des propriétés des représentations (qui seront ici des citations). 
First, \citet{Griffiths07} analyze a task where patients are asked to speak out the first word which comes to their mind when they are presented with a random alphabet letter. The authors show that there exists a link between the ease of recall of words and one of their semantic features, namely their authority position (PageRank) in a language-wide semantic network built from external word association data. % property concernant une tâche où l'on présente aux sujets une lettre de l'alphabet en leur demandant de dire le premier mot commençant par cette lettre qui leur vient à l'esprit. Les auteurs montrent qu'une certaine propriété des mots (le PageRank calculé sur un réseau sémantique construit à partir de données d'associations de mots) est un bon facteur prédictif de quel mot sera rappelé parmi les mots commençant par la lettre demandée ; ils montrent ainsi un lien entre une caractéristique sémantique quantitative des mots et la facilité de rappel de ces mots dans cette tâche. 
A second psycholinguistic study by \citet{Chan10} shows, in a picture-naming task, that words come quicker when they have a higher clustering coefficient in an underlying phonological network (which, again, is  defined from external phonological data). %, étudie une tâche où l'on demande aux sujets de nommer oralement une image. Les auteurs montrent qu'une autre propriété des mots, phonologique cette fois (le coefficient de regroupement des mots calculé sur le réseau phonologique formé par ces mots), a un effet sur la rapidité à nommer l'image présentée : c'est un autre cas de lien entre une caractéristique quantitative (phonologique) d'un mot et la facilité à produire ce mot.

### 2014-01-22 brains-protocol.tex
Phrasings and obvious mistakes

To measure cognitive bias during reformulation, we chose to focus on public representations which could be expected to undergo variations which could be described easily and automatically.
Quotations appeared to be a perfect candidate. First, they are generally cleanly delimited by quotes, which greatly facilitates their detection in text corpora. Second, they stem from a unique ``original'' version, and could ideally be traceable back to it.
Third, and perhaps most importantly, their duplication should \emph{a priori} be highly faithful, apart from cropping issues: not only should variations be of a moderate magnitude, but when specific words are not perfectly duplicated, it is safe to assume that the variation is due to involuntary cognitive bias --- as writers may expect any casual reader to easily verify, and thus criticize, the fidelity to the original quote.

@@

We thus relied on a large and reliable quotation dataset collected by \citet{Leskovec09} which 
%matched these essential criteria: made of public representations, subject to variations which could be expected to be of a relatively moderate magnitude, be ideally described by reference to an ideal original representation, and 
is of a sufficiently large size to lend itself to statistical analysis.
%\tb{On a donc choisi de s'appuyer sur un corpus déjà disponible au lieu de construire le nôtre, ce qui est une des raisons pour lesquelles l'analyse de la question est si tributaire du choix des données.  Le jeu de données utilisé par  et \citet{Simmons11} se prête bien à la notion de représentation publique qu'on a choisie ci-dessus. Construit et publié en 2009 par \citeauthor{Leskovec09}, c'est un jeu de données reconnu et utilisé dans la littérature, largement pré-traité pour nos besoins comme on le verra ci-dessous. Il est d'une dimension bien suffisante pour atteindre des niveaux de significativité intéressants, tout en restant manipulable sur un serveur de calcul modeste ou même sur un bon ordinateur portable.}
This dataset consists of the daily collection of news stories or blog posts from around a million online sources, with the approximate publication rate of 900k texts per day, over a nine-month period ranging from August 2008 to April 2009 \cite{Leskovec09-url}.\footnote{The original article~\citep{Leskovec09} does not provide additional details on the method of source selection.}
Quotations were then automatically extracted from this corpus: each quotation is a more or less faithful excerpt of an utterance (oral or written) by the cited person. Quotations were then gathered in a graph and connected according to their similarity: either because they differ by very few words (in that case, no more than one word) or because they share a certain sequence of words (in that case, at least ten consecutive words). A community detection algorithm is applied on that quotation graph to detect aggregates of tightly connected, i.e. sufficiently similar, groups of quotations (see \citet{Leskovec09} for more detail).  Eventually, the data contains around $70,000$ sets of quotations. Each set admittedly contains all variations of a same parent utterance, together with the respective URLs and timestamps of publication of each individual variation. % We have the list of occurrences, URL and publication time of the quotations of a given such set. %  which are possibly slightly differing from a parent utterance.

@@

We also considered grammatical types within quotations by detecting of so-called \emph{Part-of-Speech} (POS) categories, using the Penn TreeBank Project typology \citep{Santorini90} and thereby distinguishing verbs, nouns, adjectives and 
adverbs.  The results were however extremely similar across the various categories, exhibiting no specific behavior of words belonging to different POS categories. %. \tb{Enfin, on ajoute à ces mesures la catégorie grammaticale, dans les citations, des mots échangés lors des substitutions. On s'est restreint à quatre des catégories les plus fréquentes, en utilisant la catégorisation et la notation des \emph{Part-of-Speech} (abrégé POS)\footnote{Le terme \emph{Part-of-Speech}, ou "partie du discours" en français, désigne les catégories grammaticales des mots (ou même des parties de mots) utilisées en linguistique.} du Penn TreeBank Project\footnote{Le Penn TreeBank Project définit une typologie très complète des POS, ainsi qu'une façon de les noter : à chaque POS correspond un code de deux ou trois lettres majuscules. Par exemple, tous les POS tenant du verbe commencent par la lettre $V$ (e.g. $VBN$ pour participe passé, $VBG$ pour participe présent, etc.).}~\citep{Santorini90} : verbes (POS dont le code commence par $V$), noms (POS dont le code commence par $N$), adjectifs (POS dont le code commence par $J$), et adverbes (POS dont le code commence par $R$). Cette mesure supplémentaire affine les observables définies ci-dessus, et permet d'observer un éventuel effet des catégories grammaticales sur les résultats.}

\subsubsection{Network-based definitions}
We rely on two types of network-based data: curated synonym associations provided by WordNet~\citep{WordNet10}, and ``free association'' relationships collected by~\citep{Nelson04}.

The WordNet (WN) database gathers more than 117~000 concepts and about 147~000 words attached to these concepts. WordNet entries are disambuised words, or ``\emph{lemmas}'', which are rattached to concepts called ``synsets'', each of them gathering the set of lemmas corresponding to the same single meaning. Note that the choice of a synset for a lemma immediately determines the grammatical category of that lemma. Lemmas make the nodes of the semantic network (homonyms are aggregated into a single node), whose neighbors are synonymous lemmas belonging to the same synset(s) as the lemma(s) represented by that node.  In other words, synsets induce cliques in the semantic network. Links connecting two synonymous lemmas are being assigned a weight corresponding to the number of times these lemmas are mentioned by WordNet as synonyms, i.e. in the same synset.

@@

\bigskip
We introduce three standard network-based features, which may also be relevant psycholinguistically:
\begin{itemize}
	\item \textbf{Distance}, a classical dyadic network measure corresponding to the shortest path length between two nodes. This measure may be used as a proxy of the remoteness between two given lemmas, which is likely to have an influence on the likelihood of being a relevant substitution candidate.
	\item \textbf{Centrality} $k$, measured by the number of neighbors of a given node, and interpretable as a proxy for the polysemy of words. For the directed network FA we may also define the incoming centrality $k'$ (number of neighbors pointing \emph{to} that node).  Note that in the present case, there is a quasi-perfect correlation between node degree and node \emph{PageRank} \cite{Page99}, which had already been used in \cite{Griffiths07} and may be interpreted as a generalized and recursive measure of lemma polysemy --- central nodes in the PageRank sense are those which are linked to many nodes themselves linked to many nodes, and so on, recursively.\footnote{{Betweenness coefficient}, another measure of node centrality describing the extent to which a node tends to connect otherwise remote areas of the network \cite{free:set}. More technically, it corresponds to the normalized number of shortest paths connecting dyads which pass through that node; the higher, the more important is that node in ensuring the connectedness of the rest of the network. Correlation between FA-BC and FA-PR is $0.74$; between WN-BC and WN-PR it is $0.83$.}
	\item \textbf{Clustering coefficient} $c$, which measures the extent to which a node belongs to a local aggregate of tightly connected nodes, and defined as the ratio between the number of actual vs. possible links between a node's neighbors \cite{watt-coll}. A lemma is likely to have a higher clustering coefficient whenever the synsets which it belongs to are themselves semantically linked. This measure had been used by \citet{Chan10}.
\end{itemize}

@@

It would be difficult to discuss possible correlations between distance ($d$), on one side, and centrality ($k$ or $k'$) or clustering coefficient ($c$) on the other side, as the former is a dyadic measure while the latter are monadic measures.  We may however note that centrality and clustering are relatively weakly (and negatively) correlated, with a Pearson coefficient of $-0.26$.

An important question here relates to the various possible correlations between those variables.  The age of acquisition is a key variable, primarily as a usual suspect in psycholinguistic studies, and also appears to be generally relatively correlated to all the other variables.  This coincides with an ongoing debate suggesting that age of acquisition encodes a variety of phenomena which are difficult to disentangle from more specific phenomena, which could be captured by more independent variables such as \CN.
In effect, the minimal linear correlation of $aoa$ ($-0.31$) with other variables has a higher magnitude than the maximal linear correlation of other variables with each other ($-0.26$).

@@

Indeed, an author writing a new post and reproducing (possibly altering) a quotation bases himself on original quotations we have no access to. We face a reverse engineering problem where, given all quotations and their occurrence timestamps, we must estimate which was the originating quote for each occurrence of each quote. We therefore model the underlying quote selection process by making a few additional assumptions which let us define substitutions from the available data.

>>>>

To measure cognitive bias during reformulation, we chose to focus on public representations which could be expected to undergo variations which could be described easily and automatically.
Quotations appeared to be a perfect candidate. First, they are generally cleanly delimited by quotation marks (or by HTML markup in web pages), which greatly facilitates their detection in text corpora. Second, they stem from a unique ``original'' version, and could ideally be traceable back to it.
Third, and perhaps most importantly, their duplication should \emph{a priori} be highly faithful, apart from cropping issues: not only should variations be of a moderate magnitude, but when specific words are not perfectly duplicated, it is safe to assume that the variation is due to involuntary cognitive bias --- as writers may expect any casual reader to easily verify, and thus criticize, the fidelity to the original quote.

@@

We thus relied on a large and reliable quotation dataset collected by \citet{Leskovec09} which 
%matched these essential criteria: made of public representations, subject to variations which could be expected to be of a relatively moderate magnitude, be ideally described by reference to an ideal original representation, and 
is of a sufficiently large size to lend itself to statistical analysis.
%\tb{On a donc choisi de s'appuyer sur un corpus déjà disponible au lieu de construire le nôtre, ce qui est une des raisons pour lesquelles l'analyse de la question est si tributaire du choix des données.  Le jeu de données utilisé par  et \citet{Simmons11} se prête bien à la notion de représentation publique qu'on a choisie ci-dessus. Construit et publié en 2009 par \citeauthor{Leskovec09}, c'est un jeu de données reconnu et utilisé dans la littérature, largement pré-traité pour nos besoins comme on le verra ci-dessous. Il est d'une dimension bien suffisante pour atteindre des niveaux de significativité intéressants, tout en restant manipulable sur un serveur de calcul modeste ou même sur un bon ordinateur portable.}
This dataset consists of the daily collection of news stories or blog posts from around a million online sources, with the approximate publication rate of 900k texts per day, over a nine-month period ranging from August 2008 to April 2009 \cite{Leskovec09-url}.\footnote{The original article~\citep{Leskovec09} does not provide additional details on the method of source selection.}
Quotations were then automatically extracted from this corpus: each quotation is a more or less faithful excerpt of an utterance (oral or written) by the cited person. Quotations were then gathered in a graph and connected according to their similarity: either because they differ by very few words (in that case, no more than one word) or because they share a certain sequence of words (in that case, at least ten consecutive words). A community detection algorithm is applied to that quotation graph to detect aggregates of tightly connected, i.e. sufficiently similar, groups of quotations (see \citet{Leskovec09} for more detail).  The final data contains around $70,000$ sets of quotations. Each set allegedly contains all variations of a same parent utterance, annotated with the respective URLs and timestamps of publication of each individual variation. % We have the list of occurrences, URL and publication time of the quotations of a given such set. %  which are possibly slightly differing from a parent utterance.

@@

We also considered grammatical types within quotations by detection of \emph{Part-of-Speech} (POS) categories, using the Penn TreeBank Project typology \citep{Santorini90} and thereby distinguishing verbs, nouns, adjectives and 
adverbs.  The results were however extremely similar across the various categories, exhibiting no specific behavior of words belonging to different POS categories. %. \tb{Enfin, on ajoute à ces mesures la catégorie grammaticale, dans les citations, des mots échangés lors des substitutions. On s'est restreint à quatre des catégories les plus fréquentes, en utilisant la catégorisation et la notation des \emph{Part-of-Speech} (abrégé POS)\footnote{Le terme \emph{Part-of-Speech}, ou "partie du discours" en français, désigne les catégories grammaticales des mots (ou même des parties de mots) utilisées en linguistique.} du Penn TreeBank Project\footnote{Le Penn TreeBank Project définit une typologie très complète des POS, ainsi qu'une façon de les noter : à chaque POS correspond un code de deux ou trois lettres majuscules. Par exemple, tous les POS tenant du verbe commencent par la lettre $V$ (e.g. $VBN$ pour participe passé, $VBG$ pour participe présent, etc.).}~\citep{Santorini90} : verbes (POS dont le code commence par $V$), noms (POS dont le code commence par $N$), adjectifs (POS dont le code commence par $J$), et adverbes (POS dont le code commence par $R$). Cette mesure supplémentaire affine les observables définies ci-dessus, et permet d'observer un éventuel effet des catégories grammaticales sur les résultats.}

\subsubsection{Network-based definitions}
We rely on two types of network-based data: curated synonym associations provided by WordNet~\citep{WordNet10}, and ``free association'' relationships collected by~\citet{Nelson04}.

The WordNet (WN) database gathers more than 117~000 concepts and about 147~000 words attached to these concepts. WordNet entries are disambiguated words, or ``\emph{lemmas}'', which are attached to concepts called ``synsets'', each of them gathering the set of lemmas corresponding to the same single meaning. Note that the choice of a synset for a lemma immediately determines the grammatical category of that lemma. Lemmas make the nodes of the semantic network (homonyms are aggregated into a single node), whose neighbors are synonymous lemmas belonging to the same synset(s) as the lemma(s) represented by that node.  In other words, synsets induce cliques in the semantic network. Links connecting two synonymous lemmas are being assigned a weight corresponding to the number of times these lemmas are mentioned by WordNet as synonyms, i.e. in the same synset.

@@

\bigskip
We introduce three standard network-based features, which may also be relevant psycholinguistically:
\begin{itemize}
	\item \textbf{Distance}, a classical dyadic network measure corresponding to the shortest path length between two nodes. This measure may be used as a proxy of the remoteness between two given lemmas, which is likely to have an influence on the likelihood of being a relevant substitution candidate.
	\item \textbf{Centrality} $k$, measured by the number of neighbors of a given node, and interpretable as a proxy for the polysemy of words. For the directed network FA we also define the incoming centrality $k'$ (number of neighbors pointing \emph{to} that node).  Note that in the present case, there is a quasi-perfect correlation between node degree and node \emph{PageRank} \cite{Page99}, which had already been used in \cite{Griffiths07} and may be interpreted as a generalized and recursive measure of lemma polysemy --- central nodes in the PageRank sense are those which are linked to many nodes themselves linked to many nodes, and so on, recursively.\footnote{\TB{{Betweenness coefficient}, another measure of node centrality describing the extent to which a node tends to connect otherwise remote areas of the network \cite{free:set}. More technically, it corresponds to the normalized number of shortest paths connecting dyads which pass through that node; the higher, the more important is that node in ensuring the connectedness of the rest of the network. Correlation between FA-BC and FA-PR is $0.74$; between WN-BC and WN-PR it is $0.83$.}}
	\item \textbf{Clustering coefficient} $c$, which measures the extent to which a node belongs to a local aggregate of tightly connected nodes, and defined as the ratio between the number of actual vs. possible links between a node's neighbors \cite{watt-coll}. A lemma is likely to have a higher clustering coefficient whenever the synsets to which it belongs are themselves semantically linked. This measure had been used by \citet{Chan10}.
\end{itemize}

@@

\TB{It would be difficult to discuss possible correlations between distance ($d$), on one side, and centrality ($k$ or $k'$) or clustering coefficient ($c$) on the other side, as the former is a dyadic measure while the latter are monadic measures.  We may however note that centrality and clustering are relatively weakly (and negatively) correlated, with a Pearson coefficient of $-0.26$.} \rk{remove, redundant with below}

An important question here relates to the various possible correlations between all the variables we consider.  The age of acquisition is a key variable, primarily as a usual suspect in psycholinguistic studies, and also appears to be generally correlated to all the other variables.  This coincides with an ongoing debate suggesting that age of acquisition encodes a variety of phenomena which are difficult to disentangle from more specific phenomena, which could be captured by more independent variables such as \CN.
\TB{Indeed, the minimal linear correlation of $aoa$ ($-0.31$) with other variables has a higher magnitude than the maximal linear correlation of other variables with each other ($-0.26$).} \rk{how do we get to that conclusion?}

@@

Indeed, an author writing a new post and reproducing (possibly altering) a quotation bases himself on original quotations we have no access to. We face a reverse engineering problem where, given all quotations and their occurrence timestamps, we must estimate which was the originating quote for each instance of each quote. We therefore model the underlying quote selection process by making a few additional assumptions which let us define substitutions from the available data.

### 2014-01-22 brains-results.tex
Phrasings and obvious mistakes

We first looked at well-known (and well-studied) psycholinguistic features, and doing so we were able to determine which of the two alternative hypotheses presented in the introduction is valid; the features we examined were age of acquisition norms\footnote{The Age-of-Acquisition data is obtained from \citet{Kuperman12}.} and number of phonemes\footnote{The number of phonemes are obtained from the CMU Pronouncing Dictionary for U.S. English \citep{Weide98}, included in the NLTK package \citep{Bird09}. We note that these two measures are likely to be linked, since words with larger numbers of phonemes are likely to be learned later in development.}. First, by looking at substitution susceptibilities for those features, we can see that words with lower age of acquisition (figure \ref{fig:aoa-suscept}) and words with lower number of phonemes (figure \ref{fig:MNphonemes-suscept}) are more likely to be substituted than words with higher values for either of those features. This shows us that, although words learned later in development (as well as words with larger number of phonemes) may be cognitively harder to recall, they seem to be substituted according to hypothesis [X]: their relevance and specificness makes them less susceptible to change, whereas words learned earlier or with smaller number of phonemes are more easily substituted.

Now looking at what type of arrival words are selected upon substitution, we can see that both words learned earlier in development (figure \ref{fig:aoa-absolute}) as well as words with lower number of phonemes (figure \ref{fig:MNphonemes-absolute}) are substituted for words with roughly the same feature values (they tend to augment a little, but stay way below the values expected under $\mathcal{H}_0$ or $\mathcal{H}_{0,n}$). Conversely, words learned later in development and words with higher number of phonemes, when substituted, go towards words with lower feature values, but no lower than what the null hypotheses predict. This shows again that words learned early on are easily interchangeable for other words learned at about the same age. On the other hand, it seems the cognitive system puts little constraint on the arrival word when substituting a word learned later in development (that is, those words are rarely substituted, but when a substitution does occur the arrival word does not show constraints related to the features we examined). We also tested these results for dependence on the grammatical category of words, i.e. POS tags, and found no effect.

@@

Secondly, and coming back to our primary goal of providing a first empirical testing means for epidemiological models of culture, we added to these two features a number of more abstract properties computed from the Free Association norms used by \citet{Griffiths07}: namely, the PageRank, betweenness coefficient and clustering coefficient of the words.

>>>>

\TB{We first looked at well-known (and well-studied) psycholinguistic features, and doing so we were able to determine which of the two alternative hypotheses presented in the introduction is valid; the features we examined were age of acquisition norms\footnote{The Age-of-Acquisition data is obtained from \citet{Kuperman12}.} and number of phonemes\footnote{The number of phonemes are obtained from the CMU Pronouncing Dictionary for U.S. English \citep{Weide98}, included in the NLTK package \citep{Bird09}. We note that these two measures are likely to be linked, since words with larger numbers of phonemes are likely to be learned later in development.}.} \rk{repetition} First, by looking at substitution susceptibilities for those features, we can see that words with lower age of acquisition (figure \ref{fig:aoa-suscept}) and words with lower number of phonemes (figure \ref{fig:MNphonemes-suscept}) are more likely to be substituted than words with higher values for either of those features. This shows us that, although words learned later in development (as well as words with larger number of phonemes) may be cognitively harder to recall, they seem to be substituted according to hypothesis [X]: their relevance and specificness makes them less susceptible to change, whereas words learned earlier or with smaller number of phonemes are more easily substituted.

Now looking at what type of arrival words are selected upon substitution, we can see that both words learned earlier in development (figure \ref{fig:aoa-absolute}) as well as words with lower number of phonemes (figure \ref{fig:MNphonemes-absolute}) are substituted for words with roughly the same feature values (they tend to augment a little, but stay way below the values expected under $\mathcal{H}_0$ or $\mathcal{H}_{00}$). Conversely, words learned later in development and words with higher number of phonemes, when substituted, go towards words with lower feature values, but no lower than what the null hypotheses predict. This shows again that words learned early on are easily interchangeable for other words learned at about the same age. On the other hand, it seems the cognitive system puts little constraint on the arrival word when substituting a word learned later in development (that is, those words are rarely substituted, but when a substitution does occur the arrival word does not show constraints related to the features we examined). We also tested these results for dependence on the grammatical category of words, i.e. POS tags, and found no effect.

@@

\TB{Secondly, and coming back to our primary goal of providing a first empirical testing means for epidemiological models of culture, we added to these two features a number of more abstract properties computed from the Free Association norms used by \citet{Griffiths07}: namely, the PageRank, betweenness coefficient and clustering coefficient of the words.} \rk{repetition}
